## The HSMRF model in RevBayes

RevBayes [@Hoehna2016b] provides several methods to estimate diversification rates through time.
Here we used the [https://revbayes.github.io/tutorials/divrate/ebd.html][RevBayes tutorial] using the episodic-birth-death model with a Horseshoe-Markov-Random-Field (HSMRF) prior distribution to estimate diversification rates [@Magee2020].
We assume here that you followed the RevBayes tutorial and have the output stored as a log-file in the *output* folder.
If you have not run a RevBayes analysis and wish to explore functionality of the package, we have provided a thinned output file, accessible with `data("HSMRFBDP_primates")`.

For some of our plotting, we need the speciation times.
So we extract these from the tree.
```{r RevBayes-read-times}
btimes <- sort( as.numeric( branching.times( tree ) ) )
max_t <- max(btimes)
```

Then we read in the RevBayes log-file as a simple tab-delimited file.
```{r read-RevBayes-output, cache=TRUE}
samples <- read.table(file="output/HSMRFBDP_primates.log",
                      stringsAsFactors=FALSE,
                      header=TRUE)
```
Then, from this log-file we want to automatically extract the speciation and extinction rates.
Be careful that you used the same names as we did.
Otherwise you need to modify this code-snippet accordingly.
```{r RevBayes-extract-output, cache=TRUE}
par_speciation <- samples[, startsWith(names(samples), "speciation_rate.")]
par_extinction <- samples[, startsWith(names(samples), "extinction_rate.")]
```
In principle we can use now the full posterior estimates.
However, in this example we are only interested in the posterior median.
We strongly encourage to use the medians and not the means because the posterior median is a more robust estimate for rates, see @Magee2020.
```{r RevBayes-compute-median, cache=TRUE}
est_speciation <- apply(par_speciation, 2, median)
est_extinction <- apply(par_extinction, 2, median)
```
Since we assumed in the tutorial epochs of equal size, we can simply use the `approxfun` to create the speciation and extinction rate functions.
```{r RevBayes-functions, cache=TRUE}
lambda <- approxfun(seq(0, max_t, length.out = length(est_speciation)),
                    est_speciation)
mu <- approxfun(seq(0, max_t, length.out = length(est_extinction)),
                est_extinction)
```
Finally, let's have a look at the estimated rate functions.
```{r RevBayes-plot-functions, fig.height=4, fig.width=4, fig.align="center", fig.cap="Estimated speciation and extinction rates using RevBayes on the primates phylogeny of Springer et al. (2012)."}
ylim <- c( min(0,est_speciation,est_extinction),
           max(est_speciation,est_extinction))

curve( lambda, xlim=c(max_t, 0), lwd=2, lty=1, col="blue",
       ylim=ylim, main="Primates", xlab="time", ylab="rate",
       cex.main=1.5, cex.lab=1.5 )
curve( mu, add=TRUE, lwd=2, lty=1, col="red" )
legend("topleft",legend=c("speciation","extinction"),
       col=c("blue","red"), lty=c(1,1), cex=1.2)
```


For a later use in the tutorial where we want to compare the estimates between different software, we are going to store the rate functions.
```{r RevBayes-rates, cache=TRUE}
lambda.RevBayes <- lambda
mu.RevBayes     <- mu
```

