## TESS Analysis {#tess}

First, load the TESS package.
```{r load-TESS}
library(TESS)
```

For some of our plotting, we need the speciation times.
So we extract these from the tree.
```{r TESS-read-times}
btimes <- sort( as.numeric( branching.times( tree ) ) )
max_t <- max(btimes)
```

```{r TESS-analysis, echo=TRUE, cache=TRUE}
tess.analysis(tree,
              empiricalHyperPriors = TRUE,
              samplingProbability = 1.0,
              numExpectedRateChanges = 2,
              MAX_ITERATIONS = 100000,
              MAX_TIME = 2*60*60,
              MIN_ESS = 625,
              dir = "primates_comet")
```

```{r TESS-output, echo=TRUE, cache=TRUE}
tess.output <- tess.process.output("primates_comet",
                              numExpectedRateChanges = 2)
```

```{r TESS-compute-median, cache=TRUE}
est_speciation <- apply(tess.output[["speciation rates"]], 2, median)
est_extinction <- apply(tess.output[["extinction rates"]], 2, median)
```

Let us make a data frame representation of the rates. Since TESS is parameterized in forward time (i.e. with $t = 0$ being the root of the phylogeny, increasing toward the present), we want to reverse the rates to be compatible with `ACDC`.
```{r tess-rates-dataframe}
times <- seq(0, max_t, length.out = 100)

primates_ebd_tess <- tibble(
  "time" = times,
  "lambda" = rev(est_speciation),
  "mu" = rev(est_extinction),
)
head(primates_ebd_tess)
```
Now let's have a look at the estimated rate functions
```{r plot-tess-rates, fig.width=5, fig.height = 5, fig.cap = "The estimated posterior median results from an episodic birth-death model fitted in RevBayes."}
p <- primates_ebd_tess %>%
  tidyr::gather("rate", "value", -time) %>%
  ggplot(aes(x = time, y = value, color = rate)) +
  geom_line() +
  theme_classic() +
  scale_x_reverse() +
  theme(legend.position = c(0.2, 0.2)) +
  labs(y = "rate", x = "time before present")
plot(p)
```

The data frame `primates_ebd_tess` is what we get when we call the `data("primates_ebd_tess")` function in `ACDC`. We use the RevBayes estimated rates in our vignette examples, but it is easy to substitute those for the TESS analysis results to construct the congruence class.